---
title: "Georgia Birth Weight Modeling"
author: "Alejandro Hernandez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
# clear workspace
rm(list = ls())

# load relevant libraries
library(tidyverse)
library(knitr)
library(rigr)
library(geepack)
library(lme4)
library(nlme)


# color selection
colors <- c("#FC600A", # dark orange
            "#C21460", # dark pink
            "#3F0000") # darker red
```

```{r data-loading}
birthwt <- read.csv("data/cdc birthwt.csv")

names(birthwt) <- c("mother.id", "birth.order", "birth.weight", "maternal.age", "child.id")

birthwt <- birthwt %>%
  # create birth weight binary factor
  mutate(birth.weight.binary = ifelse(birth.weight < 2500, 1, 0)) %>%
  # create a interpregnancy variable
  arrange(mother.id, birth.order) %>%
  group_by(mother.id) %>%
  mutate(interval = maternal.age - lag(maternal.age)) %>%
  ungroup

birthwt %>% filter(interval < 0)
# mothers 57939 and 212988 both have negative intervals for their third births
# we will remove all their data and do not expect measurable changes in future 
# analysis, their data is not special among the sample
birthwt <- birthwt %>% filter(!mother.id %in% c(57939, 212988))
```

### Modeling correlation

```{r model-correlation, echo=FALSE}
# produce a spaghetti plot of birth weight over maternal age
birthwt %>%
  ggplot(aes(y = birth.weight, x = maternal.age)) +
    geom_line(aes(group = mother.id), alpha = 0.3, color = colors[1]) + 
    geom_smooth(method = "loess", se = F, color = colors[3], lwd =  0.75) +
    geom_hline(yintercept = 2500, color = "blue") +
    geom_hline(yintercept = 1500, color = "red") +
    xlab("Maternal age (yr)") + 
    ylab("Birth weight (g)") +
    theme_bw()

# produce a spaghetti plot of birth weight over birth order
birthwt %>%
  ggplot(aes(y = birth.weight, x = birth.order)) +
    geom_line(aes(group = mother.id), alpha = 0.3, color = colors[1]) + 
    geom_point(data = . %>% group_by(birth.order) %>% 
                 summarize(mean = mean(birth.weight)),
               aes(x = birth.order, y = mean), size = 2.5, color = colors[3]) +
    geom_hline(yintercept = 2500, color = "blue") +
    geom_hline(yintercept = 1500, color = "red") +
    xlab("Birth order") + 
    ylab("Birth weight (g)") +
    theme_bw()

# Autocorrelation function
acf(birthwt$birth.weight)
```

We presume that average population effects on birth weight will be well-modeled by a generalized estimating equation (GEE) model with a auto-regressive correlation structure, based on the decaying pattern of the ACF graph above.

For the same reason, we expect the individual-specific effects on birth weight to be well-modeled by a linear mixed-effects model (LMM) that also assumes an auto-regressive correlation structure.

We elect to fit many LMM and GEE models, assuming various correlation structures. We expect to report coefficients and performance diagnostics for four models:

-   

    (1) LMM with no covariates, random intercept and fixed slope, and no assumed correlation structure.

-   

    (3) LMM with covariate for birth order, random intercept and fixed slope, and assumed auto-regressive correlation structure.

-   

    (4) GEE with no predictors and auto-regressive working correlation structure.

-   

    (5) GEE with covariate for birth order and auto-regressive working correlation structure.

### Modeling birth weight

```{r lmm-models, error=TRUE}
#### Linear Mixed-Effects Models ####
# Random intercept, fixed slope; No covariates
lmm0 <- nlme::lme(birth.weight ~ maternal.age, 
            random = ~ 1 | mother.id, 
            data = birthwt)

# Random intercept, fixed slope; with birth order covariate
lmm1 <- lme(birth.weight ~ maternal.age + birth.order,
            random = ~ 1 | mother.id, 
            data = birthwt)

# Random intercept and slope; with birth order covariate
lmm2 <- lme(birth.weight ~ maternal.age + birth.order,
            random = ~ maternal.age | mother.id, 
            data = birthwt)

# Random intercept and fixed slope; with birth order covariate and auto-regressive correlation structure
lmm3 <- lme(birth.weight ~ maternal.age + birth.order, 
            random = ~ 1 | mother.id, correlation = corAR1(), 
            data = birthwt)

# Random intercept and slope; with birth order covariate and auto-regressive correlation structure
# WILL NOT COMPLETE; DOES NOT CONVERGE
lmm4 <- lme(birth.weight ~ maternal.age + birth.order, 
            random = ~ maternal.age | mother.id, correlation = corAR1(), 
            data = birthwt)

# Random intercept and slope; with birth order covariate and exchangeable 
# correlation structure
lmm5 <- lme(birth.weight ~ maternal.age + birth.order, 
            random = ~ maternal.age | mother.id, correlation = corCompSymm(), 
            data = birthwt)

# Random intercept and fixed slope; with birth order covariate and exponential 
# correlation structure
lmm6 <- lme(birth.weight ~ maternal.age + birth.order, 
            random = ~ 1 | mother.id, correlation = corExp(), 
            data = birthwt)

# Random intercept and slope; with birth order covariate and exponential 
# correlation structure
# WILL NOT COMPLETE; DOES NOT CONVERGE
lmm7 <- lme(birth.weight ~ maternal.age + birth.order, 
            random = ~ maternal.age | mother.id, correlation = corExp(), 
            data = birthwt)
```

```{r gee-models}
#### Generalized Estimating Equations ####

# Auto-regressive correlation structures assume that the correlation between
# observations decreases as the time between them increases
gee0 <- geepack::geeglm(birth.weight ~ maternal.age, id = mother.id,
               corstr = "ar1", data = birthwt)

gee1 <- geeglm(birth.weight ~ maternal.age + birth.order, id = mother.id, 
               corstr = "ar1", data = birthwt)

# Exchangeable correlation structures assume that all pairs of observations 
# within a subject have the same correlation, regardless of the time interval 
# between them
gee2 <- geeglm(birth.weight ~ maternal.age, id = mother.id,
               corstr = "exchangeable", data = birthwt)

gee3 <- geeglm(birth.weight ~ maternal.age + birth.order, id = mother.id, 
               corstr = "exchangeable", data = birthwt)

# Independent correlation structures assume that all measurements within a 
# subject are uncorrelated
gee4 <- geeglm(birth.weight ~ maternal.age, id = mother.id,
               corstr = "independence", data = birthwt)

gee5 <- geeglm(birth.weight ~ maternal.age + birth.order, id = mother.id, 
               corstr = "independence", data = birthwt)
```

# Diagnostics

```{r lmm-diagnostics, echo=FALSE}
## Diagnostics for LMM models of birth weight
# exclude models 4 and 7, they did not converge

# AIC (relative goodness of fit)
print("AIC scores, a measure of relative goodness of fit")
AIC(lmm0, lmm1, lmm2, lmm3, lmm5, lmm6) %>% arrange(AIC)

# Variance and correlation components
print("Variance/correlation of LMM Model 0")
lmm0 %>% VarCorr
print("Variance/correlation of LMM Model 1")
lmm1 %>% VarCorr
print("Variance/correlation of LMM Model 2")
lmm2 %>% VarCorr
print("Variance/correlation of LMM Model 3")
lmm3 %>% VarCorr
print("Variance/correlation of LMM Model 5")
lmm5 %>% VarCorr
print("Variance/correlation of LMM Model 6")
lmm6 %>% VarCorr

# Residual plot
ggplot(mapping = aes(y = residuals(lmm0), x = fitted(lmm0))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "LMM 0")

ggplot(mapping = aes(y = residuals(lmm1), x = fitted(lmm1))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "LMM 1")

ggplot(mapping = aes(y = residuals(lmm2), x = fitted(lmm2))) +
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "LMM 2")

ggplot(mapping = aes(y = residuals(lmm3), x = fitted(lmm3))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "LMM 3")

ggplot(mapping = aes(y = residuals(lmm5), x = fitted(lmm5))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "LMM 5")

ggplot(mapping = aes(y = residuals(lmm6), x = fitted(lmm6))) +
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "LMM 6")

# QQ plot
qqnorm(residuals(lmm0))
qqline(residuals(lmm0))
title(sub="Plot for LMM model 0")

qqnorm(residuals(lmm1))
qqline(residuals(lmm1))
title(sub="Plot for LMM model 1")

qqnorm(residuals(lmm2))
qqline(residuals(lmm2))
title(sub="Plot for LMM model 2")

qqnorm(residuals(lmm3))
qqline(residuals(lmm3))
title(sub="Plot for LMM model 3")

qqnorm(residuals(lmm5))
qqline(residuals(lmm5))
title(sub="Plot for LMM model 5")

qqnorm(residuals(lmm6))
qqline(residuals(lmm6))
title(sub="Plot for LMM model 6")
```

```{r gee-diagnostics, echo=FALSE}
## Diagnostics for GEE models of birth weight

# Residual plot
ggplot(mapping = aes(y = residuals(gee0), x = fitted(gee0))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "GEE 0")

ggplot(mapping = aes(y = residuals(gee1), x = fitted(gee1))) +
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "GEE 1")

ggplot(mapping = aes(y = residuals(gee2), x = fitted(gee2))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "GEE 2")

ggplot(mapping = aes(y = residuals(gee3), x = fitted(gee3))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "GEE 3")

ggplot(mapping = aes(y = residuals(gee4), x = fitted(gee4))) +
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "GEE 4")

ggplot(mapping = aes(y = residuals(gee5), x = fitted(gee5))) + 
  geom_point(alpha = 0.2) + geom_smooth() + labs(title = "GEE 5")

# QQ plot
qqnorm(residuals(gee0))
qqline(residuals(gee0))
title(sub="Plot for GEE model 0")

qqnorm(residuals(gee1))
qqline(residuals(gee1))
title(sub="Plot for GEE model 1")

qqnorm(residuals(gee2))
qqline(residuals(gee2))
title(sub="Plot for GEE model 2")

qqnorm(residuals(gee3))
qqline(residuals(gee3))
title(sub="Plot for GEE model 3")

qqnorm(residuals(gee4))
qqline(residuals(gee4))
title(sub="Plot for GEE model 4")

qqnorm(residuals(gee5))
qqline(residuals(gee5))
title(sub="Plot for GEE model 5")
```

# Results

```{r lmm-results, echo=FALSE}
# Summary for all results
# lmm0 %>% summary %>% coef
# lmm1 %>% summary %>% coef
# lmm2 %>% summary %>% coef
# lmm3 %>% summary %>% coef
# lmm5 %>% summary %>% coef
# lmm6 %>% summary %>% coef

# Table of reported results
# Report LMMs 0 and 3
lmm0_res <- lmm0 %>% summary %>% coef %>% data.frame %>% 
  mutate(CI.lower = Value - Std.Error, CI.upper = Value + Std.Error)

lmm3_res <- lmm3 %>% summary %>% coef %>% data.frame %>%
  mutate(CI.lower = Value - Std.Error, CI.upper = Value + Std.Error)

lmm_res <- rbind(lmm0_res, lmm3_res)[-c(1,3),] # ignore intercept coefficients
lmm_res$Model <- c("Mixed", "Mixed + Birth order + Autoreg", "")
lmm_res$Effect <- c("Maternal age", "Maternal age", "Birth order")
# Rename coefficients
lmm_res %>% 
  select(Model, Effect, Estimate = Value, CI.lower, CI.upper) %>%
  knitr::kable(row.names = FALSE, digits = 3, 
               caption = "Linear mixed-effects model results")
```

All the LMM models agreed with one another, except for Model 0 (first row of table above) which estimated a weaker effect from maternal age than the others.

Recall,

```         
# Random intercept, fixed slope; No covariates
lmm0 <- nlme::lme(birth.weight ~ maternal.age, 
            random = ~ 1 | mother.id, 
            data = birthwt)

# Random intercept and fixed slope; with birth order covariate and auto-regressive
# correlation structure
lmm3 <- lme(birth.weight ~ maternal.age + birth.order, 
            random = ~ 1 | mother.id, correlation = corAR1(), 
            data = birthwt)
```

```{r gee-results, echo=FALSE}
# Summary for all results
# gee0 %>% summary %>% coef
# gee1 %>% summary %>% coef
# gee2 %>% summary %>% coef
# gee3 %>% summary %>% coef
# gee4 %>% summary %>% coef
# gee5 %>% summary %>% coef

# Table of reported results
# Report GEEs 0 and 1
gee0_res <- gee0 %>% summary %>% coef %>% data.frame %>% 
  mutate(CI.lower = Estimate + Std.err, CI.upper = Estimate + Std.err)

gee1_res <- gee1 %>% summary %>% coef %>% data.frame %>%
  mutate(CI.lower = Estimate + Std.err, CI.upper = Estimate + Std.err)


gee_res <- rbind(gee0_res, gee1_res)[-c(1,3),] # ignore intercept coefficients
gee_res$Model <- c("Autoregressive", "Autoregressive + Birth order", "")
gee_res$Effect <- c("Maternal age", "Maternal age", "Birth order")
gee_res %>% 
  select(Model, Effect, Estimate, CI.lower, CI.upper) %>%
  knitr::kable(row.names = FALSE, digits = 3, 
               caption = "Generalized estimating equations results")
```

All the GEE models agree that adding a covariate increases the effects of age (by about 7 units). However, the other GEE models estimate different effects (I think they'll all higher). The ACF plot in the beginning convinced me that assuming auto-regressive and temporal decay of correlation is something we should stick to.

\pagebreak

# Modeling log odds of low birth weight

NOTE I AM ASSUMING THIS IS THE CORRECT WAY TO INTERPRET GLMM'S AS SIMILAR TO LOGISTIC REGRESSION; I REALIZE NOW I MAY HAVE BEEN WRONG PLEASE VERIFY

```{r}
# Generalized linear mixed-effects models (GLMM) to model log odds of low birth
# weight 
glmm0 <- lme4::glmer(
  birth.weight.binary ~ maternal.age + (1 | mother.id),
  data = birthwt %>% na.omit, family = binomial)

glmm1 <- glmer(
  birth.weight.binary ~ maternal.age + interval + (1 | mother.id),
  data = birthwt %>% na.omit, family = binomial)

glmm2 <- glmer(
  birth.weight.binary ~ maternal.age + birth.order + (1 | mother.id),
  data = birthwt %>% na.omit, family = binomial)

glmm3 <- glmer(
  birth.weight.binary ~ maternal.age + interval + birth.order + (1 | mother.id),
  data = birthwt %>% na.omit, family = binomial)

# Summary for all results
glmm0_res <- glmm0 %>% summary %>% coef %>% data.frame %>%
    mutate(CI.lower = Estimate + Std..Error, CI.upper = Estimate + Std..Error)

glmm1_res <- glmm1 %>% summary %>% coef %>% data.frame %>%
    mutate(CI.lower = Estimate + Std..Error, CI.upper = Estimate + Std..Error)

glmm2_res <- glmm2 %>% summary %>% coef %>% data.frame %>%
    mutate(CI.lower = Estimate + Std..Error, CI.upper = Estimate + Std..Error)

glmm3_res <- glmm3 %>% summary %>% coef %>% data.frame %>%
    mutate(CI.lower = Estimate + Std..Error, CI.upper = Estimate + Std..Error)

# Table for all results
gee_res <- rbind(glmm0_res, glmm1_res, glmm2_res, glmm3_res)[-c(1,3,6,9),] %>%
  exp %>% select("exp(Estimate)" = Estimate, CI.lower, CI.upper)
gee_res$Model <- c("A", 
                   "B", "",
                   "C", "",
                   "D", "", "")
gee_res$Effect <- c(
  "Maternal age",
  "Maternal age", "Interval",
  "Maternal age", "Birth order",
  "Maternal age", "Interval", "Birth order")

gee_res %>% 
  select(Model, Effect, everything()) %>%
  knitr::kable(row.names = FALSE, digits = 3, 
               caption = "Generalized linear mixed-effects results")
```

Exp(Estimates) is on the odds scale, please verify that if an effect's 95% CI covers 1, the we cannot conclude it has significant effect.

I don't know how to compare models through diagnostics yet.

**End of document.**
